Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-05 analysis/paper/paper.qmd
minor edits, move China spheroids to front of paper

# Context of Qianshangying and its spheroids

@@

ggplot(df_long, aes(x = material, y = prop, fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    title = "Raw Material Proportions in Spheroids, Cores, and Total",
    x = "Raw Material",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  scale_fill_manual(values = c(
    "spheroids" = "orange", 
    "cores" = "#66c2a5", 
    "total" = "lightgray"
  )) +
  theme_minimal() +
  theme(
    legend.position = c(0.95, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black")  
  )

@@

# 画图
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Type"
  ) +
  scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "lightgray"
  )) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = c(0.95, 0.95),       
    legend.justification = c("right", "top"),  
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA), 
    legend.box.background = element_rect(color = "black")  
  )

@@

In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size. Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively. The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids. Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts. @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant. @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items. Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.\@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250). Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced. Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM. We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

@@

To further investigate technological strategies, we incorporated orientation statistics as a complementary approach. This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts. We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were carried out in Geomagic Wrap, Rhino8 and R.

@@

knitr::kable(data_raw, caption = "Summary of core typology and morphological parameters.")

@@

We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the QSY assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

@@

```{r, echo=FALSE}
#| label: fig-spheroids-2797-1098
#| fig-cap: "Diacritical analysis of QSY spheroids"

@@

According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size). This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.

Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the QSY facted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].

The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation]. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013]. Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015]. Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product. <!-- BM: this is well written and an appropriate discussion, but I think we should be prepared for some peer reviewers to get excited and try to defend the Acheulean as the very first time shaping strategies are evident! Probably because they have published a paper on that topic and don't want to be proved wrong -->

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of QSY facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relyingon morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at QSY, our quantitative analyses clearly differentiate polyhedrons from spheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at QSY represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at QSY, indicating a diversity of reduction goals.

However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores. In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns. At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable. However, it remains unclear whether this typological transformation was intentional or unplanned. In addition, the spheroids at QSY appear to have undergone another form of typological transformation—from hammerstones to spheroids. One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.

## Technological and typological variability of spheroids in China

<!-- BM: this is an excellent section, but seems out of sequence here. I think we should split it up. Put the first few paragaphs at the start of the paper as background to spheroids in the region. Then put the last few para and the bits that directly mention QSY here in the discussion to give more context to the importance of the results from QSY. What do you think?-->

file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
data3 <- read_excel(file_path)

site data: 1. Mujiaqiao, [@Wei1984]; 2. Liaohe river region, [@Li1991]; 3. Deshan Second Brick factory, [@Xi1994]; 4. Zhangjiatan, [@LixianMuseum1992]; 5. Huzhua Mount., [@Tan1999]; 6. Fangniu Mount., [@Fang2002]; 7. Fanba, [@Xia2022Fanba]; 8. Xiaokongshan, [@XiaokongshanTeam1988]; 9. Hejialiang, [@Wang2014Hanzhong]; 10. Longgangsi, [@Lu2006Liangshan]; 11. Shangdan basin, [@Wang2013Danjiang]; 12. Yaoshi basin, [@Wang2000Yaoshi]; 13. Choushuihe, [@Dai1973Lantian]; 14. Kehe, [@Jia1962Kehe]; 15. Shuigou, [@Huang1964Sanmenxia]; 16. Hezhigou, [@Liu1984Jingchuan]; 17. Xigou, [@Jia1959Xigou]; 18. Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19. Liujiacha, [@Xie1982Liujiacha]; 20. Dangcheng, [@Wu1989Dangcheng] ; 21. Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22. Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23. Jinsitai, [@Wang2010Jinstai]

```{r, echo=FALSE}
#| label: fig-Spheroids_distribution_China
#| fig-cap: "Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries"

In China, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids_distribution_China. In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene. In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai]. The widespread occurrence of spheroids across time and space may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China. However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites. However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids_distribution_China). Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology. Rather, what we infer is that faceted spheroids in China represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another. If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, this would suggest that shaping techniques originated in an even earlier technological context.

This pattern aligns with spheroid findings in other regions of the world. Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan. Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence]. The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019]. Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020]. In particular, faceted spheroids, which reflect a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.

@@

This study provides one of the earliest systematic analyses focused on spheroids from northern China. By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the QSY spheroids. Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion. The evidence of shaping strategies on QSY spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores. This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages. Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries. However, the QSY assemblage from \~429ka clearly reveals a higher level of technological sophistication.

This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China. We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology. Some assemblages contain spheroids that may be comparable to those from QSY and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than QSY. This highlights the particular significance of the QSY spheroids. However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors. We therefore emphasize the need for site-specific analysis of spheroids in future research.

@@

```{R}
?pie
```

>>>>

## Spheroids in East Asia

file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
data3 <- readxl::read_excel(file_path)

::: {#fig-Spheroids-distribution-China}

Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries.  1. Mujiaqiao, [@Wei1984]; 2. Liaohe river region, [@Li1991]; 3. Deshan Second Brick factory, [@Xi1994]; 4. Zhangjiatan, [@LixianMuseum1992]; 5. Huzhua Mount., [@Tan1999]; 6. Fangniu Mount., [@Fang2002]; 7. Fanba, [@Xia2022Fanba]; 8. Xiaokongshan, [@XiaokongshanTeam1988]; 9. Hejialiang, [@Wang2014Hanzhong]; 10. Longgangsi, [@Lu2006Liangshan]; 11. Shangdan basin, [@Wang2013Danjiang]; 12. Yaoshi basin, [@Wang2000Yaoshi]; 13. Choushuihe, [@Dai1973Lantian]; 14. Kehe, [@Jia1962Kehe]; 15. Shuigou, [@Huang1964Sanmenxia]; 16. Hezhigou, [@Liu1984Jingchuan]; 17. Xigou, [@Jia1959Xigou]; 18. Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19. Liujiacha, [@Xie1982Liujiacha]; 20. Dangcheng, [@Wu1989Dangcheng] ; 21. Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22. Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23. Jinsitai, [@Wang2010Jinstai]

:::

In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China. In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene. In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai]. The widespread occurrence of spheroids across time and space in East Asia may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites. However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids-distribution-China). Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology. Rather, what we infer is that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another. If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, then this would suggest that shaping techniques may have originated in an even earlier technological context.

This pattern aligns with spheroid findings in other regions of the world. Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan. Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence]. The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019]. Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020]. In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.

# Qianshangying and its spheroids

@@

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Core",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total"
           )), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Total" = "lightgray"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.95, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black")  
  )

@@

# 画图
p_dimensions <- 
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Technological Type"
  ) +
  scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "lightgray"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.95, 0.95),       
    legend.justification = c("right", "top"),  
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA), 
    legend.box.background = element_rect(color = "black")  
  )

@@

In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics. Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively. The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids. Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

```{r}
#| label: fig-basic-characteristics
#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing of core, spheroid, and multifacial core dimensions"

library(cowplot)

plot_grid(p_raw_material,
          p_dimensions,
          labels = "AUTO")

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts. @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant. @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items. Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons. @Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250). Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced. Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM. We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

@@

To further investigate technological strategies, we incorporated orientation statistics as a complementary approach. This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts. We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were carried out in Geomagic Wrap, Rhino8 and R 4.5.1.

@@

knitr::kable(data_raw %>% 
               # round values for display
               mutate(across(where(is.numeric), round, 3)), 
             caption = "Summary of core typology and morphological parameters.")

@@

We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

@@

```{r, echo=FALSE}
#| label: fig-spheroids-2797-1098
#| fig-cap: "Diacritical analysis of Qianshangying spheroids"

@@

According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids from Qianshangying. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@@fig-basic-characteristics). This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.

Therefore, it can be inferred that the Qianshangying hominins deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the Qianshangying facetted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].

The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation]. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013]. Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015]. Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product. <!-- BM: this is well written and an appropriate discussion, but I think we should be prepared for some peer reviewers to get excited and try to defend the Acheulean as the very first time shaping strategies are evident! Probably because they have published a paper on that topic and don't want to be proved wrong -->

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of the Qianshangying facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.

However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores. In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns. At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable. However, it remains unclear whether this typological transformation was intentional or unplanned. In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids. One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.

@@

This study provides one of the earliest systematic analyses focused on spheroids from East Asia. By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids. Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion. The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores. This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages. Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries. However, the Qianshangying assemblage from \~429ka clearly reveals a higher level of technological complexity

This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China. We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology. Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying This highlights the particular significance of the Qianshangying spheroids. However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors. We therefore emphasize the need for site-specific analysis of spheroids in future research.

@@
